Reanalysis Step 4: Main Effects Outcome Analysis

### 2025-08-27 reanalysis-4-outcome_analysis.qmd
caching and emmeans

library(mice)
library(survey)
library(broom)
library(marginaleffects)

@@

```{r fit-models}
# Store fitted models and extracted coefficients
fitted_models <- list()
model_summaries <- list()

for (i in 1:n_imputations) {
  # Get dataset and weights
  imp_data <- imputed_datasets[[i]]
  imp_data$weights <- twang_weights[[i]]

@@

@@

```{r marginal-effects}
# Calculate AME for each imputed dataset then pool
ame_list <- list()

for (i in 1:n_imputations) {
  imp_data <- imputed_datasets[[i]]
  imp_data$weights <- twang_weights[[i]]

@@

@@

```{r predicted-probs}
# Calculate predicted probabilities for different scenarios
# Using the first imputation for illustration
imp_data <- imputed_datasets[[1]]
imp_data$weights <- twang_weights[[1]]

# Create prediction scenarios
scenarios <- expand.grid(
  ever_lapse_binary = c(0, 1),
  dr_severity = c("No_DR", "NPDR", "PDR"),
  any_treatment = c(0, 1)
)

# Set other variables to their weighted means/modes
weighted_means <- svydesign(ids = ~1, weights = ~weights, data = imp_data) |>
  svymean(~baseline_VA_logMAR + age + CCI + DCSI, na.rm = TRUE)

# Get modes for categorical variables
get_mode <- function(x, w) {
  tbl <- xtabs(w ~ x)
  names(tbl)[which.max(tbl)]
}

scenarios$baseline_VA_logMAR <- weighted_means["baseline_VA_logMAR"]
scenarios$age_cat <- get_mode(imp_data$age_cat, imp_data$weights)
scenarios$gender_cat <- get_mode(imp_data$gender_cat, imp_data$weights)
scenarios$race_ethnic_cat <- get_mode(imp_data$race_ethnic_cat, imp_data$weights)
scenarios$insurance_cat <- get_mode(imp_data$insurance_cat, imp_data$weights)
scenarios$CCI <- round(weighted_means["CCI"])
scenarios$DCSI <- round(weighted_means["DCSI"])
scenarios$glaucoma_bef_hitplus_cat <- 0
scenarios$otherretina_bef_hitplus_cat <- 0
scenarios$catsurg_before_hitplus_cat <- 0

# Get predictions
predictions <- predict(fitted_models[[1]], newdata = scenarios, type = "response")

# Format results
pred_table <- scenarios[, c("ever_lapse_binary", "dr_severity", "any_treatment")]
pred_table$Predicted_Prob <- round(predictions, 3)
pred_table$Lapse <- ifelse(pred_table$ever_lapse_binary == 1, "Lapsed", "Not Lapsed")
pred_table$Treatment <- ifelse(pred_table$any_treatment == 1, "Treated", "Not Treated")

# Reshape for better display
library(tidyr)
pred_wide <- pred_table |>
  select(Lapse, dr_severity, Treatment, Predicted_Prob) |>
  pivot_wider(names_from = Lapse, values_from = Predicted_Prob) |>
  arrange(dr_severity, Treatment)

kable(pred_wide, 
      caption = "Predicted probability of VI/blindness by lapse status",
      col.names = c("DR Severity", "Treatment", "Not Lapsed", "Lapsed"))

# Calculate risk differences
pred_wide$Risk_Difference <- pred_wide$Lapsed - pred_wide$`Not Lapsed`
pred_wide$Relative_Risk <- pred_wide$Lapsed / pred_wide$`Not Lapsed`

kable(pred_wide |> select(dr_severity, Treatment, Risk_Difference, Relative_Risk), 
      caption = "Risk differences and relative risks",
      col.names = c("DR Severity", "Treatment", "Risk Difference", "Relative Risk"),
      digits = 3)
```

>>>>

library(mice)
library(survey)
library(broom)
library(marginaleffects)
library(emmeans)

@@

```{r fit-models}
# Check if models are already cached
models_cache_file <- file.path(reanalysis_data_dir, "fitted_models_main_effects.rds")
force_refit <- FALSE  # Set to TRUE if you want to force re-fitting

if (file.exists(models_cache_file) && !force_refit) {
  # Load cached models
  cached_models <- readRDS(models_cache_file)
  fitted_models <- cached_models$fitted_models
  model_summaries <- cached_models$model_summaries
  cat("Loaded cached models from:", models_cache_file, "\n")
  cat("To refit models, set force_refit <- TRUE\n")
} else {
  # Fit new models
  cat("Fitting models to each imputed dataset...\n")

  # Store fitted models and extracted coefficients
  fitted_models <- list()
  model_summaries <- list()

  for (i in 1:n_imputations) {
    cat("Fitting model", i, "of", n_imputations, "\r")

    # Get dataset and weights
    imp_data <- imputed_datasets[[i]]
    imp_data$weights <- twang_weights[[i]]

@@

  cat("\n")

  # Save to cache
  cached_models <- list(
    fitted_models = fitted_models,
    model_summaries = model_summaries,
    formula_used = main_effects_formula,
    n_imputations = n_imputations,
    timestamp = Sys.time()
  )

  saveRDS(cached_models, models_cache_file)
  cat("Models cached to:", models_cache_file, "\n")
}

@@

```{r marginal-effects}
# Check if AMEs are cached
ame_cache_file <- file.path(reanalysis_data_dir, "ame_results_main_effects.rds")
force_recalc_ame <- FALSE  # Set to TRUE to force recalculation

if (file.exists(ame_cache_file) && !force_recalc_ame) {
  # Load cached AMEs
  ame_cache <- readRDS(ame_cache_file)
  ame_list <- ame_cache$ame_list
  cat("Loaded cached AMEs from:", ame_cache_file, "\n")
} else {
  # Calculate AME for each imputed dataset then pool
  cat("Calculating AMEs for each imputed dataset...\n")
  ame_list <- list()

  for (i in 1:n_imputations) {
    cat("Calculating AME", i, "of", n_imputations, "\r")
    imp_data <- imputed_datasets[[i]]
    imp_data$weights <- twang_weights[[i]]

@@

  cat("\n")

  # Save to cache
  ame_cache <- list(
    ame_list = ame_list,
    n_imputations = n_imputations,
    timestamp = Sys.time()
  )
  saveRDS(ame_cache, ame_cache_file)
  cat("AMEs cached to:", ame_cache_file, "\n")
}

@@

```{r predicted-probs}
# Use emmeans for predicted probabilities - it handles survey models properly
# We'll use the first imputed model for illustration

# Get estimated marginal means for key combinations
emm_results <- tryCatch({
  # Get predictions for lapse x dr_severity x any_treatment
  emmeans(fitted_models[[1]], 
          ~ ever_lapse_binary * dr_severity * any_treatment,
          type = "response",  # Get probabilities not log-odds
          weights = "proportional")  # Use observed proportions for other variables
}, error = function(e) {
  # If emmeans fails, try a simpler specification
  emmeans(fitted_models[[1]], 
          ~ ever_lapse_binary + dr_severity + any_treatment,
          type = "response",
          weights = "proportional")
})

# Convert to data frame for display
if (!is.null(emm_results)) {
  pred_df <- as.data.frame(emm_results)

  # Clean up column names
  if ("prob" %in% names(pred_df)) {
    pred_df$Predicted_Prob <- round(pred_df$prob, 3)
  } else if ("response" %in% names(pred_df)) {
    pred_df$Predicted_Prob <- round(pred_df$response, 3)
  } else if ("emmean" %in% names(pred_df)) {
    # If we're on the link scale, transform to probability
    pred_df$Predicted_Prob <- round(plogis(pred_df$emmean), 3)
  }

  # Create lapse and treatment labels
  pred_df$Lapse <- ifelse(pred_df$ever_lapse_binary == 1, "Lapsed", "Not Lapsed")
  pred_df$Treatment <- ifelse(pred_df$any_treatment == 1, "Treated", "Not Treated")

  # Select and reshape for display
  library(tidyr)
  pred_wide <- pred_df |>
    select(Lapse, dr_severity, Treatment, Predicted_Prob) |>
    pivot_wider(names_from = Lapse, values_from = Predicted_Prob) |>
    arrange(dr_severity, Treatment)

  kable(pred_wide, 
        caption = "Predicted probability of VI/blindness by lapse status (from first imputation)",
        col.names = c("DR Severity", "Treatment", "Not Lapsed", "Lapsed"),
        digits = 3)

  # Calculate contrasts for risk differences
  # This properly accounts for the survey design
  contrasts_lapse <- tryCatch({
    pairs(emmeans(fitted_models[[1]], 
                  ~ ever_lapse_binary | dr_severity * any_treatment,
                  type = "response"))
  }, error = function(e) {
    NULL
  })

  if (!is.null(contrasts_lapse)) {
    contrasts_df <- as.data.frame(contrasts_lapse)
    kable(contrasts_df |> 
          select(dr_severity, any_treatment, estimate, SE, p.value) |>
          mutate(
            Treatment = ifelse(any_treatment == 1, "Treated", "Not Treated"),
            Risk_Difference = round(estimate, 3),
            SE = round(SE, 4),
            p.value = format.pval(p.value, digits = 4, eps = 0.0001)
          ) |>
          select(dr_severity, Treatment, Risk_Difference, SE, p.value),
          caption = "Risk differences (Lapsed - Not Lapsed) with proper variance estimation",
          col.names = c("DR Severity", "Treatment", "Risk Diff", "SE", "P-value"))
  }
} else {
  cat("Note: Predicted probabilities using emmeans could not be calculated.\n")
  cat("The main analysis results above remain valid.\n")
}
```
